# Multiple linear regression {#sec-multiple-linear-regression}

```{r}
#| include: false

source("_common.R")
```

::: {.chapterintro data-latex=""}
In most practical settings, more than one explanatory variable is likely to be associated with a response. This chapter discusses how the ideas behind simple linear regression can be extended to a model with multiple predictor variables. 

There are several applications of multiple regression. One of the most common applications in a clinical setting is estimating an association between a response variable and primary predictor of interest while adjusting for possible confounding variables. 

<!--

Fix cross refs
Sections~\ref{introductionMultipleLinearRegression} and \ref{simpleVsMultipleRegression} introduce the multiple regression model by examining the possible association between cognitive function and the use of statins after adjusting for potential confounders. Section~\ref{modelSelection} discusses another application of multiple regression\textemdash constructing a model that effectively explains the observed variation in the response variable. 
-->

The other sections in the chapter outline general principles of multiple regression, including the statistical model, methods for assessing quality of model fit, categorical predictors with more than two levels, interaction, and the connection between ANOVA and regression.  The methods used to conduct hypothesis tests and construct confidence intervals for regression coefficients extend naturally from simple to multiple linear regression, so the section on the statistical model for multiple regression can be treated as optional.
:::

```{r}
#| include: false
terms_chp_1 <- c("data")
```

