@tbl-crabsVarDescriptions provides the definitions of the variables. It is important to check the definitions of variables, as they are not always clear. In the `crabs` data, the definitions specify that the measurements of `width` and `weight` are in metric units and the variable `spine` records the condition of the two spines near the abdomen of each crab.

```{r}
#| label:  tbl-crabsVarDescriptions
#| tbl-cap:  "Variable definitions for horseshoe crab data"
#| tbl-pos: H

crabs_var_name <- c("color",
                     "spine",
                     "width",
                     "satell",
                     "weight")
crabs_var_description <- c("color of the carapace",
                            "condition of the spines along the edge of the abdomen",
                            "width of the carapace in cm",
                            "number of male satellites",
                            "weight in kg")
crabs_var_table  <- data.frame(crabs_var_name, crabs_var_description)
crabs_var_table  <- crabs_var_table  |> 
          rename('Variable name' = crabs_var_name,
                  'Variable description' = crabs_var_description)

crabs_var_table  |> 
  kbl(linesep = "", booktabs = TRUE, align = "cl")|>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    latex_options = c("striped"),
    full_width = FALSE)
```



```{r}
#| label:  tbl-pdx-parks-df
#| tbl-cap:  "The first 3 and last trees in the Portland tree data"
#| tbl-pos: H

portland_park_trees |> 
  mutate(rn = row_number()) |> 
  slice(1:3, max(rn)) |>
  relocate(rn) |> 
  rename(Case = rn) |> 
  kbl(linesep = "", booktabs = TRUE, align = "rrr", digits = 3) |>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    latex_options = c("striped"),
    full_width = FALSE)
```


Like relative frequency tables for the distribution a single categorical variable, contingency tables can be converted to show proportions. @tbl-wvs-edu-health-joint-prop show the proportion of responses for each cell in the table. The number $0.148$ in row `High`, column `Good` is the proportion of respondents with `Education_level` `High` and `Good` `Health_status`.  The values in the cells apart from the row and column totals are called elements of the joint distribution for education and health, since they specify the proportion of times each combination of values occur.  The Total column 


```{r wvs-edu-health-joint-proportion}
#| label: tbl-wvs-edu-health-joint-prop
#| tbl-cap:  "Health rating by educational level, with joint proportions"
#| tbl-pos: H
wvs_hlth_edu_cc |>
 count(Health_status, Education_level) |> 
 mutate(proportion = n / sum(n)) |>
 select(-n)  |> 
 pivot_wider(names_from = Health_status, 
              values_from = proportion) |> 
 relocate("Very Good", .after = "Education_level")  |> 
 arrange(replace(row_number(), 1, n() + 1)) |> 
 adorn_totals(where = c("row", "col")) |> 
 kbl(linesep = "", booktabs = TRUE, digits = 3) |> 
 kable_styling(
    bootstrap_options = c("striped", "condensed"),
    latex_options = c("striped"),
    full_width = FALSE) |>
 add_header_above(c(" " = 1, "Health_status" = 5, " " = 1)) |>
 column_spec(1, width = "8em") |>
 column_spec(2:6, width = "4em")  

```



```{r}
#| label: fig-leap-rel-freq-bar-plot
#| fig-cap: |
#|   Two bar plots displaying outcome by intervention in LEAP.
#| fig-subcap:
#|   - frequency bar plot showing counts
#|   - relative frequency bar plot showing proportions
#| fig-alt: |
#|   Two bar plots (frequency and relative frequency) displaying outcome to the olfactory challenge test
#|   The number of failures within each intervention is small, but the proportion of failures
#|   in the avoidance group is more than 7 times as larger than in the consumption group.
#| fig-width: 4
#| layout-ncol: 2
 LEAP_analyzed <- LEAP |> 
  dplyr::filter(stratum == "SPT-Negative") |> 
  dplyr::filter(!is.na(overall.V60.outcome))  |> 
  mutate(Outcome = case_when(
      overall.V60.outcome == "PASS OFC" ~ "Pass OFC",
      overall.V60.outcome == "FAIL OFC" ~ "Fail OFC"))  |> 
  mutate(intervention = case_when(
    treatment.group == "Peanut Avoidance" ~ "Avoidance",
    treatment.group == "Peanut Consumption" ~ "Consumption"))

ggplot(LEAP_analyzed, aes(x = intervention, fill = Outcome)) +
    scale_fill_manual(values = c(IMSCOL["red", "full"], IMSCOL["blue", "full"])) +
    geom_bar() +
    labs(x = "Intervention", y = "Count")

ggplot(LEAP_analyzed, aes(x = intervention, fill = Outcome)) +
   scale_fill_manual(values = c(IMSCOL["red", "full"], IMSCOL["blue", "full"])) +
   geom_bar(position = "fill") +
   labs(x = "Intervention", y = "Proportion") 

```
```{r crabs-color-spine-sat-box}
#| label: fig-crabs-color-spine-satell-boxplot
#| fig-cap: |
#|   Two side-by-side box plots of the distribution number of male satellites.
#| fig-subcap:
#|   - Side-by-side box plot of number of satellites by female color
#|   - Side-by-side box plot of number of satellites by female spine condition
#| fig-alt: | 
#|   The median number of male satellites shows a strong relationship with female color,
#|    with numbers of satellites decreasing as color darkens. There is no evident relationship with 
#     worsening spine condition.
#| fig-asp: 0.4
#| 

ggplot(crabs_recoded, aes(x = color, y = satell)) +
  geom_boxplot() +
  scale_color_openintro("two") +
  labs(y = "Number of satellites", x = "Color") +
  scale_y_continuous(breaks = seq(0, 16, 4))

ggplot(crabs_recoded, aes(x = spine, y = satell)) +
  geom_boxplot() +
  scale_color_openintro("two") +
  labs(y = "Number of satellites", x = "Spine condition") +
  scale_y_continuous(breaks = seq(0, 16, 4))

```
[`Portland Tree Inventory project`](https://www.portland.gov/trees/get-involved/treeinventory) is a comprehensive inventory of trees in Portland, Oregon. The project records many variables for each tree, including the date the tree was added to the inventory, its common name,  the financial value of the tree and it value as a carbon storage organism.  


```{r}
library(pdxTrees)
set.seed(080546)
#pull data from specific parks using custom get_pdxTrees_parks() function 
pdx <- get_pdxTrees_parks(park = c("Laurelhurst Park", 
                                   "Kenilworth Park",
                                   "Gov Tom McCall Waterfront Park", 
                                   "Columbia Childrens Arboretum"))
portland_trees  <- pdx  |> 
                select(Inventory_Date, Common_Name, Tree_Height,
                      Structural_Value, Carbon_Storage_lb, Park)  |> 
                sample_n(6)
```

Table @tbl-portland-trees shows a few of these variables for a random selection of trees from 4 Portland Parks. ^[Laurelhurst Park, Kenilworth Park, Gove Tom
McCall Waterfront Park and the Columbia Childrens Arboretum], and table @tbl-portland-trees-var-defs shows the definitions of these variables.  (The full dataset is available in the R package [`pdxTrees`](https://cran.r-project.org/web/packages/pdxTrees/index.html).)


```{r}
#| label:  tbl-portland-trees
#| tbl-cap:  "A random selection of 6 trees in the Portland tree dataset"
#| tbl-pos: H

portland_trees |> 
  kbl(linesep = "", booktabs = TRUE, align = "rrr")|>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    latex_options = c("striped"),
    full_width = FALSE)
```

```{r}
pdx_tree_var_name = c("Inventory_Date", "Common_Name", "Tree_Height", 
                       "Structural_Value", "Carbon_Storage_lb",
                       "Park")
pdx_tree_var_description = c("Date of data collection",
                              "Common name of the tree",
                              "Height from the ground to the live top of the tree, measured in feet. 
                              For dead trees, total height was measured.",
                              "Monetary value of replacing the tree and the benefits that 
                              it provides, based on methods from the Council of Tree and                                     Landscape Appraisers",
                              "The amount of carbon (in lbs.) that is bound up in both 
                              the above-ground and below-ground parts of the tree",
                              "The park where the tree is located")
pdx_var_table <- data.frame(pdx_tree_var_name, pdx_tree_var_description)
pdx_var_table <- pdx_var_table |> 
  rename("Variable name" = pdx_tree_var_name,
         "Variable description" = pdx_tree_var_description)
```
```{r}
#| label:  tbl-portland-trees-var-defs
#| tbl-cap:  "Variables and their descriptions for the Portland tree dataset"
#| tbl-pos: H

pdx_var_table  |> 
  kbl(linesep = "", booktabs = TRUE, align = "ll")|>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    latex_options = c("striped"),
    full_width = FALSE)
```

Even relationships between just two variables can take many forms. [The World Bank Data Group](https://data.worldbank.org/) provides free and open access to indicators of economic and social development in countries around the world. The data `wdi_2022` in the [openintro](http://openintrostat.github.io/openintro) R package contains data for some of those indicators for the year 2022. @fig-wdi_2022-income-infant-mortality shows a \index{scatterplot} **scatterplot**


```{r}
#| label: fig-wdi_2022-income-infant-mortality
#| fig-cap: A scatterplot of infant mortality rate (on the y-axis) versus the 
#|    per-capita income for 217 countries as of 2022.  Infant mortality rate
#|    is measured as infant deaths before 1 year of age among 1,000 live births.
#| fig-alt: slightly modified text here

ggplot(wdi_2022, aes(x = gni_percap, y = infant_mortality_rate)) +
  geom_point(alpha = 0.3, fill = IMSCOL["black", "full"], shape = 21) +
  labs(
    x = "Income per capita",
    y = "Infant mortality rate"
  ) +
  geom_point(
    data = wdi_2022 |> 
    dplyr::filter(country == "Argentina"),
    size = 3, stroke = 2, color = IMSCOL["red", "full"], shape = 1
  ) +
  geom_text(
    data = wdi_2022 |> 
    dplyr::filter(country == "Argentina"),
    label = "Argentina", fontface = "italic",
    nudge_x = 21, nudge_y = -5, color = IMSCOL["red", "full"]
  ) +
  guides(color = FALSE) +
  geom_segment(
    data = wdi_2022 |> 
    dplyr::filter(country == "Argentina"),
    aes(
      x = 0, y = infant_mortality_rate, xend = gni_percap, 
      yend = infant_mortality_rate,
      color = IMSCOL["red", "full"]
    ), linetype = "dashed"
  ) +
  geom_segment(
    data = wdi_2022 |> 
    dplyr::filter(country == "Argentina"),
    aes(
      x = gni_percap, y = 0, xend = gni_percap, yend = infant_mortality_rate,
      color = IMSCOL["red", "full"]
    ), linetype = "dashed"
  ) +
  scale_x_continuous(labels = waiver()) +
  scale_y_continuous(labels = waiver())
```

```{r dds-discr-box-hist}
#| label: fig-dds-discr-boxplot-histogram
#| fig-cap: Box plot and histogram of the distribution of `expenditures` in `dds.discr`
#| fig-alt: |
#|   Two plots showing the distribution of same variable, Expenditures. The upper image is a 
#|   box plot, the lower is histogram. 
#| fig-asp: 0.6
#| layout-ncol: 1
#| 

library(patchwork)
p1 <- ggplot(dds.discr, 
       aes(x = expenditures)) +
  geom_boxplot(outlier.size = 2.5) +
  theme(
    axis.title.y = element_blank(),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank() 
      ) +
  labs(x = NULL, y = NULL) +
  scale_x_continuous(limits = c(0, 80000))


p2 <- ggplot(dds.discr,
       aes(x = expenditures)) +
  labs(x = "Expenditures (USD)", 
       y = "Count") +
  geom_histogram(binwidth = 5000, 
                 closed = "left",
                 position = "dodge",
                 breaks = seq(0, 80000, 5000)) +
  scale_y_continuous(limits = c(0,500))
  
  
p1/p2





```

@tbl-dds-discr-DF} shows the first five rows of the dataset, and the variables are described in @tbl-dds-discr-variables}.



```{r dds-discr-df}
#| label:  tbl-dds-discr-DF
#| tbl-cap:  "First 5 rows of the DDS data matrix"
#| tbl-pos: H

dds.discr |>
  slice(1:5) |> 
  kbl(linesep = "", booktabs = TRUE, align = "lrrlrl")|>
  kable_styling(
    bootstrap_options = c("striped", "condensed"),
    latex_options = c("striped"),
    full_width = FALSE)
```



Move to exercises

In the scatterplot in fig-dds-age-exp-hist(a),  `expenditure` (vertical axis) is plotted against the continuous variable `age` (horizontal axis). It shows the relationship between age and expenditures evident in @fig-dds-age-boxplot, but in more detail.   The smooth line added to the plot shows the same gradual increase in expenditures until approximately age  20, with a sharp bend in the curve between roughly age 21 and 25. After age 25 expenditures... 

```{r dds-age-exp-scatter-hist}
#| label: fig-dds-age-exp-scatter-hist
#| fig-cap: |
#|       The relationship between `expenditures` and `age` and the distribution of 
#|       `expenditures`, restricted to `Hispanics` and 
#|          `White non-Hispanics`
#| sub-cap: |
#|    - Scatterplot of `expenditures` versus `age`
#|    - Histogram of `expenditures`
#| 
#| alt-cap: |
#|       Coming
#| fig-width: 8
#| layout-ncol: 2
#| 
dds_reformat |> 
  filter(ethnicity == "Hispanic" | ethnicity == "White non-Hispanic") |> 
  ggplot(aes(x = age, y = expenditures)) +
  geom_point(alpha = 0.3,  
             shape = 21, 
             fill = IMSCOL["black", "full"],
             position = position_jitter(width = 0.5, height = 0.6)) +
  geom_smooth(method = "loess", se = FALSE, color = IMSCOL["blue", "full"]) +
  labs(
    x = "Age",
    y = "Expenditures")

dds_reformat |> 
  filter(ethnicity == "Hispanic" | ethnicity == "White non-Hispanic") |> 
  ggplot(aes(x = expenditures)) +
  labs(x = "Expenditures (USD)", 
       y = "Count") +
  geom_histogram(binwidth = 5000, 
                 closed = "left",
                 position = "dodge",
                 breaks = seq(0, 80000, 5000)) +
  scale_y_continuous(limits = c(0,500))
  
  
  

```



```{r dds-eth-dodged-bar}
#| label: dds-ethnicity-dodged-bar
#| fig-cap: |
#|       A dodged bar plot showing the relationship between ethnicity and age
#| alt-cap: |
#|       Coming
#| fig-width: 8
dds_reformat |> 
  filter(ethnicity == "Hispanic" | ethnicity == "White non-Hispanic") |> 
  ggplot(aes(x = age.cohort, fill = ethnicity)) +
  geom_bar(position = "dodge") +
  labs(x = "Age category",
       y = "Count",
       fill = "Ethnicity" ) 

```







