
*Discrete random variables*

<!--- new --->

**The minimum of 2 dice.** Suppose a random process consists of rolling two dice. Using the information in @tbl-sum-two-dice find the distribution of the minimum value of the two faces.



<!--- OS4 3.29 -->

**College Smokers.**  At a university, 13% of 
students smoke.

  a. Calculate the expected number of smokers in a random sample of 100 students 
from this university.
  
  b. The university gym opens at 9 am on Saturday mornings. One Saturday morning 
at 8:55 am there are 27 students outside the gym waiting for it to open. Should 
you use the same approach from part (a) to calculate the expected number of 
smokers among these 27 students?

<!--- OS4 3.30 -->

**Ace of clubs wins.**  Consider the following card game with a well-shuffled deck of cards. If you draw a red card, you win nothing. If 
you get a spade, you win \$5. For any club, you win \$10 plus an extra \$20 for 
the ace of clubs.

  a. Create a probability model for the amount you win at this game. Also, find 
the expected winnings for a single game and the standard deviation of the 
winnings.

  b. What is the maximum amount you would be willing to pay to play this game? 
Explain your reasoning.

<!--- OS4 3.31 --->

**Hearts win.** In a new card game, you start
with a well-shuffled full deck and draw 3 cards without replacement.
If you draw 3 hearts, you win \$50. If you draw 3 black cards, you win \$25. For any other draws, you 
win nothing.

  a. Create a probability model for the amount you win at this game, and find 
the expected winnings. Also compute the standard deviation of this distribution.

  b. If the game costs \$5 to play, what would be the expected value and 
standard deviation of the net profit (or loss)? \textit{(Hint: 
profit = winnings $-$ cost; $X-5$)}

  c. If the game costs \$5 to play, should you play this game? Explain.
  
<!--- OS4 3.34 --->

**Baggage fees.**  An airline charges the following 
baggage fees: \$25 for the first bag and \$35 for the second. Suppose 54\% of 
passengers have no checked luggage, 34\% have one piece of checked luggage and 
12\% have two pieces. We suppose a negligible portion of people check more than 
two bags.

  a. Build a probability model, compute the average revenue per passenger, and 
compute the corresponding standard deviation.

  b. About how much revenue should the airline expect for a flight of 120 
passengers? With what standard deviation? Note any assumptions you make and if 
you think they are justified.

<!--- oibiostat 3.5 --->

**Gull clutch size.**  Large black-tailed gulls usually lay one to three eggs, and rarely have a fourth egg clutch. It is thought that clutch sizes are effectively limited by how effectively parents can incubate their eggs. Suppose that on average, gulls have a 25\% of laying 1 egg, 40\% of laying 2 eggs, 30\% chance of laying 3 eggs, and 5\% chance of laying 4 eggs.

  a. Calculate the expected number of eggs laid by a random sample of 100 gulls. 
		
  b. Calculate the standard deviation of the number of eggs laid by a random sample of 100 gulls.
  
  
<!--- OS4  3.35 --->

**American roulette** The game of American  roulette involves spinning a wheel with 38 slots: 18 red, 18 black, and 2 green. 
A ball is spun onto the wheel and will eventually land in a slot, where each slot 
has an equal chance of capturing the ball. Gamblers can place bets on red or 
black. If the ball lands on their color, they double their money. If it lands on 
another color, they lose their money. Suppose you bet \$1 on red. What's the 
expected value and standard deviation of your winnings?

<!--- OS4 3.36 -->

**European roulette** The game of European 
roulette involves spinning a wheel with 37 slots: 18 red, 18 black, and 1 green. 
A ball is spun onto the wheel and will eventually land in a slot, where each slot 
has an equal chance of capturing the ball. Gamblers can place bets on red or 
black. If the ball lands on their color, they double their money. If it lands on 
another color, they lose their money.

  a. Suppose you play roulette and bet \$3 on a single round. What is the 
expected value and standard deviation of your total winnings?

  b. Suppose you bet \$1 in three different rounds. What is the expected value 
and standard deviation of your total winnings?

  c. How do your answers to parts (a) and (b) compare? What does this say about 
the riskiness of the two games?

  
*Continuous random variables*  

<!--- OS4 3.37 --->

**Cat weights**  The histogram shown below represents 
the weights (in kg) of 47 female and 97 male cats. ^[W. N. Venables and B. D. Ripley. 
Modern Applied Statistics with S, Fourth edition. \url{www.stats.ox.ac.uk/pub/MASS4}. New York: Springer, 2002.] 

  a. What fraction of these cats weigh less than 2.5 kg?

  b. What fraction of these cats weigh between 2.5 and 2.75 kg?
  
  d. What fraction of these cats weigh between 2.75 and 3.5 kg?

```{r}
library(openintro)

# load MASS for data ------------------------------------------------
library(MASS)
data(cats)

# histogram of weights ----------------------------------------------

hist(cats$Bwt, breaks = seq(2, 4, 0.25), ylim = c(0, 35), 
     xlab = "Body weight",
     ylab = "Count",
     col = COL[1], main = "", 
     axes = FALSE)
axis(1)
axis(2, at = seq(0,40,10))

```
<!--- add distributions from NHANESA -->



```{r}
library(ggplot2)

library(nhanesA)
# browseNHANES() opens nhanes site in browser
#  BXO_L is from 2021 - 2023 post-pandemic cycle
bpx  <- nhanes("BPXO_L")   # Body Measures
demo <- nhanes("DEMO_L")  # Demographics

bpx_demo <- inner_join(bpx, demo, by = "SEQN")

bp <- bpx_demo |> 
  filter(!is.na(BPXOSY1),
         !is.na(BPXODI1)) |> 
  select(BPXOSY1, BPXODI1, RIAGENDR, RIDAGEYR) 


# demographics not used for now but may be useful
n_cases <- nrow(bp)
mean_sbp_bp <- round(mean(bp$BPXOSY1), 2)
med_sbp_bp <- round(median(bp$BPXOSY1), 2)
mean_dbp_bp <- round(mean(bp$BPXODI1), 2)
med_dbp_bp <- round(median(bp$BPXODI1), 2)

```

<!--- new problem --->

**Systolic blood pressure**

Systolic blood pressure (SBP) is the maximum pressure blood exerts against the walls of arteries when the heart is beating.  It is measured in mm Hg (mercury) when it is measured with a blood pressure cuff.  @fig-systolic-bp-prop is a histogram showing the distribution of SBP in `r n_cases` from the 2021 - 2023 cycle of the NHANES health survey.

  a.  Describe the distribution.
  
  b. How would you expect the mean and median to compare?
  
  c. Approximately what proportion of SBP measurements fall below 90mm HG?
  
  d. Approximately what proportion of SPB measurements fall between 90 and 175mm HG?
  
  e. If the histogram had been drawn on a density scale, would the shape of the curve change?
  
  f. If the histogram had been drawn on a density scale, how would the vertical axis labels change.
  
```{r}
#| label: fig-systolic-bp-prop
#| out-width: 80%
#| fig-cap: | 
#|     Distribution systolic blood pressure from NHANES.
ggplot(bp, aes(x = BPXOSY1)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
         breaks = seq(60, 220, 5),
         closed = "right") +
     labs(x = "Systolic blood pressure", y = "Proportion") +
     scale_x_continuous(
       breaks = seq(60, 220, 10))+
    scale_y_continuous(breaks = seq(0, .15, 0.01))

```
<!--- new problem --->

**Diastolic blood pressure**

Diastolic blood pressure (DBP) is the pressure blood exerts against the walls of arteries when the heart is beating resting and is also measured in mm HG.   @fig-systolic-bp-prop is a histogram showing the distribution of DBP in `r n_cases` from the 2021 - 2023 cycle of the NHANES health survey.

  a. Describe the distribution.
  
  b. How would you expect the mean and median to compare?
  
  c. Approximately what proportion of DBP measurements fall below 55mm HG?
  
  d. Approximately what proportion of DPB measurements fall between 60 and 70mm HG?
  
```{r}
#| label: fig-diastolic-bp-prop
#| out-width: 80%
#| fig-cap: | 
#|     Distribution diastolic blood pressure from NHANES.
ggplot(bp, aes(x = BPXODI1)) +
  geom_histogram(aes(y = after_stat(count / sum(count))),
         breaks = seq(30, 120, 5),
         closed = "right") +
     labs(x = "Diastolic blood pressure", y = "Proportion") +
     scale_x_continuous(
       breaks = seq(30, 120, 10))+
    scale_y_continuous(breaks = seq(0, .20, 0.01))

```



```{r}
#  BMX_J is 2021 - 2023 post-pandemic cycle
bmx  <- nhanes("BMX_L")   # Body Measures
demo <- nhanes("DEMO_L")  # Demographics

bmx_demo <- inner_join(bmx, demo, by = "SEQN")

youth <- bmx_demo |> 
  filter(RIDAGEYR < 18,
         !is.na(BMXWT),
         !is.na(BMXHT),
         !is.na(RIAGENDR),
         BMXWT > 0,
         BMXHT > 0) |> 
  select(BMXWT, BMXHT, RIAGENDR) 

n_cases_youth <- nrow(youth)
mean_weight <- round(mean(youth$BMXWT), 2)
med_weight <- round(median(youth$BMXWT), 2)

```

<!--- new problem --->

**Weights of US youths**

@fig-youth-wt-density is a density plot of the weight in kg  of `r n_cases_youth` participants in the NHANES 2021 - 2023 survey who were less than 18 years of age.

  a.  Describe the distribution.
  
  b.  What is the probability that a randomly chosen individual from this group weighs less than 25kg?
  
  c.  What is the probability that someone chosen randomly weights between 25 and 35 kg?
  
  d.  What other variable associated with weight might be helpful to know to understand the shape of the plot?
  
  e.  What proportion of this sample weighs 10kg?

```{r}
#| label: fig-youth-wt-density
#| out-width: 80%
#| fig-cap: | 
#|     Distribution weight (kg) for people under age 18, from NHANES.

ggplot(youth, aes(x = BMXWT)) +
  geom_histogram(aes(y = after_stat(density)),
                 breaks = seq(0, 150, 5),
                 closed = "right") +
  labs(x = "Weight (kg)", y = "Density") +
  scale_x_continuous(
    breaks = seq(0, 150, 25)) +
  scale_y_continuous(
    breaks = seq(0, 0.3, 0.005)
  )

```

*Sums of independent random variables*

<!--- OS4 3.44 --->

**Scooping ice cream.** Ice cream usually comes in 1.5 
quart boxes (48 fluid ounces), and ice cream scoops hold about 2 ounces. 
However, there is some variability in the amount of ice cream in a box as well as 
the amount of ice cream scooped out. We represent the amount of ice cream in the 
box as $X$ and the amount scooped out as $Y$. Suppose these random variables have 
the following means, standard deviations, and variances:

| Variable | Mean | SD   | Variance |
|:----------|-----:|----:|---------:|
| $X$       | 48   | 1   | 1.0000   |
| $Y$       | 2    | 0.25| 0.0625   |

  a. An entire box of ice cream, plus 3 scoops from a second box is served at a 
party. How much ice cream do you expect to have been served at this party? What 
is the standard deviation of the amount of ice cream served?

  b. How much ice cream would you expect to be left in the box after scooping 
out one scoop of ice cream? That is, find the expected value of $X-Y$. What is 
the standard deviation of the amount left in the box?

  c. Using the context of this exercise, explain why we add variances when we 
subtract one random variable from another.

<!--- new problem --->

**The cost of health care, I**  Suppose the HMO described in @sec-cost-of-hmo-insurance is considering doubling the fee for both the premium and for each visit.

   a.  What will be the the expected cost for each member?
   
   b.  What will be the variance?
   
**The coast of health care, II**  Two individuals who share a large house decide to create a common budget.  They have the same insurance with the health plan described in @sec-cost-of-hmo-insurance and both are between 25 and 40 and have no chronic diseases.

  a.  What is their expected total annual cost of health care?
  
  b.  What is the variance of their total cost?  What assumption did you make and is it reasonable?

  c.  Is the variance in part b the same as that calculated in the previous problem?  Explain.

<!--- OS4 3.45 --->

**Variance of a mean, I**  Suppose we have independent observations $X_1$ and $X_2$ froma distribution with mean $\mu$ and standard deviation $\sigma$.
What is the variance of the mean of the two values:
$\frac{X_1 + X_2}{2}$?

<!--- OS4 3.46 --->

**Variance of a mean, II.**  Suppose we have 3 independent observations
$X_1$, $X_2$, $X_3$ from
a distribution with mean $\mu$ and standard deviation $\sigma$.
What is the variance of the mean of these 3 values:
$\frac{X_1 + X_2 + X_3}{3}$?

<!---OS4 3.47 --->

**Variance of a mean, III.**  Suppose we have $n$ independent observations
$X_1$, $X_2$, ..., $X_n$ from
a distribution with mean $\mu$ and standard deviation $\sigma$.
What is the variance of the mean of these $n$ values:
$\frac{X_1 + X_2 + \dots + X_n}{n}$?

*Distributions of pairs of random variables*

**Joint distributions.**

Consider two random variables, $X$ and $Y$, with the joint distribution shown in @tbl-joint-dist-exercise. 

|        | $Y = 1$ | $Y = 2$ |
|:------:|:------:|:------:|
| $X = 1$ | 0.20 | 0.40 |
| $X = 4$ | 0.30 | 0.10 |

: Joint distribution of $X$ and $Y$. {#tbl-joint-dist-exercise}

  a.  Compute the marginal distributions of $X$ and $Y$. 

  b.  Identify the joint probability $p_{X,Y}(1, 2)$. 

  c.  What is the value of $p_{X, Y}(2, 1)$? 

  d.  Compute the conditional distribution of $X$ given that $Y = 2$.

<!--- new problem--->

**Marginal means and standard deviations.**

Compute the mean and standard deviation for $X$ and $Y$ with joint distribution @tbl-joint-dist-exercise.

<!--- new problem -->
**Checking independence.**

Are the random variables with joint distribution in @tbl-joint-dist-exercise independent?

<!--- new problem --->
**Correlated of random variables.**

  a.  Calculate the correlation of the variables $X$ and $Y$ with joint distribution in @tbl-joint-dist-exercise, using @eq-spec-form-corr-alt.
  
  b. Use @tbl-joint-dist-exercise to explain why $X$ and $Y$ are negatively correlated.
  
<!--- new problem --->

**Variance of a sum of correlated variables.**

  a. Using @eq-general-var-sum-rvs Compute the variance and standard deviation of $X + Y$ if $X$ and $Y$ have the joint distribution in @tbl-joint-dist-gprac.    
  
  b. Compare the variance of $X + Y$ with $\operatorname{Var}(X) + \operatorname{Var}(Y)$ and explain the difference.
  
  c. Compute the variance and standard deviation of $X - Y$.
